Hybrid oddball - SSVEP BCI

### 2013-06-28 draftHybridPaper/draftHybridBCI.tex
been a while

\acrodef{SSVEP}{Steady-State Visually Evoked Potential}
\acrodef{BCI}{Brain-Computer Interface}
\acrodef{ERP}{Event Related Potential}
\acrodef{SVM}{Support Vector Machine}
\acrodef{EEG}{electroencephalography}

@@

\begin{abstract}
Objectives:
blablabla blalabbla \ac{SSVEP}-based BCIs.

@@

% P3 BCI, definition, plus and minus
Some of the earliest \ac{EEG}-\ac{BCI} systems were based on the P3 component of the \ac{ERP} (\cite{Farwell1988, Donchin2000}).
The P3 is a positive deflection in the EEG time-locked to salient stimuli presented in an oddball paradigm, typically evoked over the parietal cortex, and occurs between \SIlist[list-units = single]{200;500}{\ms} after stimulus onset \cite{Sutton1965}.
Although those \acp{BCI} rely mostly on the P3 component, other components (\emph{e.g.}, occipital N1 and/or N200) may also be used for ERP detection \cite{Bianchi2010, Kaufmann2011}, for this reason we prefer here to use the term \emph{oddball-based \acp{BCI}}.

@@

% SSVEP BCI, definition, plus and minus
Other systems of interest are \acp{BCI} based on \acp{SSVEP}.
They rely on the psychophysiological properties of the EEG brain responses recorded from the occipital cortex during the periodic presentation of identical visual stimuli (\emph{i.e.} flickering stimuli).
When the periodic presentation is at a sufficiently high rate (\SI{>6}{\Hz}), stable and synchronized neural oscillations at the stimulus frequency and its harmonics are evoked over the visual cortex \cite{Regan1966, Herrmann2001, Luck2005}.
Such \acp{BCI} are particularly attractive because \acp{SSVEP} have high signal-to-noise ratios and are less susceptible to eye movement and blink artifacts \cite{Perlstein2003} as well as electromyographic artifacts \cite{Gray2003}.
Several \ac{SSVEP}-based \acp{BCI} have been successfully tested with healthy subjects (see \cite{Vialatte2010} for a review).

@@

    % Stimuli presentation
    All stimulation employed MATLAB\textsuperscript{\textregistered}, the stimuli were visually presented on a laptop's LCD screen (\SI{60}{\Hz} refresh rate) and their display and timing used the \emph{Psychophysics Toolbox Extensions} (\cite{Brainard1997,Pelli1997}).

@@

        % exp description
        As shown in \autoref{fig:stimSeq}, a typical \emph{stimulation cycle}, starts with a \SI{2000}{\ms} cue, indicating the participant his/her target item, followed by a \SI{1000}{\ms} pause during which the cue disappears and all icons remained gray.
        The background rectangle starts then to flicker and the oddball stimulation begins \SI{500}{\ms} later.
        The oddball stimulation consisted of 10~\emph{flashing sequences} during which each of the 6~icons was flashed one after another in random order for a duration randomly set between \SIlist[list-units = single]{200;300}{\ms}.
        As usually done for oddball experiments, the participants were instructed to focus on their target symbol and count the number of time it flashes.
        A \SI{1000}{\ms} pause followed the oddball stimulation and preceded the next cue.
        An \emph{experimental run} lasted approximately 4~minutes and consisted of 12 consecutive stimulation cycles, so that each of the 6~icons was cued twice (in random order).

        As we aimed here at studying the effect of the flickering background on the oddball \ac{ERP} response, we considered 5~experimental conditions.
        The first one (baseline) consisted of a run as described in the previous paragraph but in which no flickering background was displayed.
        The 4~other conditions differed only by the frequency of the flickering background; the frequencies used were \SIlist[list-units = single]{8.57;10;12;15}{\Hz}, corresponding to the division of the refreshing rate of the screen by \numlist{7;6;5;4}, respectively.

@@

        % use
        The data collected were used to compare the shape of the oddball response (response to the flashing of the target icon) and the \ac{ERP} classification accuracy (response to target \emph{v.s.} response to non-target flasing) across conditions.

@@

        % use
        This experimental design allows a comparison of the \ac{SSVEP} responses of the participants with and without an oddball stimulation superimposed on the \ac{SSVEP} stimulation for a different set of \ac{SSVEP} frequencies.

@@

        After studying individually oddball \acp{ERP} and \ac{SSVEP} responses, we aim here at studying the possibility to detect both responses

@@

        \subsubsection{Experiment 1: ERPs classification}
        \label{sec:2.3.1AnalysisExp1}

        % preprocessing: filtering

        % studying the shape of ERP: epoch rejection

%%%%%%%%        \subsubsection{Data Analysis: \acs{ERP} classification}
%%%%%%%%        \label{sec:2.2.3AnalysisErpClassification}

        % studying classification accuracy: no epoch rejection
        We are interested in studying the effect of the flickering stimulus on the target \ac{ERP} detection accuracy.
        As the accuracy of an oddball \ac{BCI} is known to depend on the number of repetitions of flashing sequences used to average the \acp{ERP}; the classification accuracy was measured on \acp{ERP} averaged over the $n$~first repetitions of the flashing sequence with $n$~going from 1~to~10.
        This was done so as to mimic a real use of the system (where the stimulation sequence would stop after $n$~repetitions) and so that the accuracies are always measured on the same number of detection attempts.

        For each experimental condition

        Data were filtered between \SIlist[list-units = single]{0.3;15}{\Hz} (zero-phase 3\rd order Butterworth filter), cut into \SI{800}{\ms} epochs starting from the stimuli onsets and downsampled to \SI{128}{\Hz}.

        For each trial (stimulus), we had 8~channels~$\times$~80~data~points~=~640~features to classify as a response to either a target stimulus or a non-target stimulus.
        A linear \emph{Support Vector Machine} (SVM; \cite{Cristianini2000,Suykens2002}) with a 10-fold cross-validation and a linesearch for the optimization of the regularization parameter was built from the normalized training features.
        Training the linear SVM with the modified finite Newton method proposed by \textcite{Keerthi2006} took around one minute.

@@

    \subsection{Experiment 1: studying the oddball \acsp{ERP}}
    \label{sec:2.2Oddball}

    \subsection{Experiment 2: studying the \acs{SSVEP} responses}
    \label{sec:2.3SSVEP}

    \subsection{Experiment 3: hybrid classification}
    \label{sec:2.4Hybrid}

>>>>

% fixes the issue with labels due to the acronym package
% http://tex.stackexchange.com/questions/109178/why-does-the-compiler-keeps-telling-me-forever-to-rerun-because-labels-have-ch
\makeatletter
\newcommand{\extraclearlabels}{\protected@write\@auxout{}{%
  \string\reset@newl@bel
}}
\makeatother

% define acronyms
\acrodef{SSVEP}{Steady-State Visually Evoked Potential}
\acrodef{BCI}{Brain-Computer Interface}
\acrodef{ERP}{Event Related Potential}
\acrodef{SVM}{Support Vector Machine}
\acrodef{EEG}{electroencephalography}

@@

\begin{abstract}
\extraclearlabels
Objectives:
blablabla blalabbla \ac{SSVEP}-based BCIs.

@@

% P3 BCI, definition, plus and minus
Some of the earliest \ac{EEG}-\ac{BCI} systems were based on the P3 component of the \ac{ERP} \parencite{Farwell1988, Donchin2000}.
The P3 is a positive deflection in the EEG time-locked to salient stimuli presented in an oddball paradigm, typically evoked over the parietal cortex, and occurs between \SIlist[list-units = single]{200;500}{\ms} after stimulus onset \parencite{Sutton1965}.
Although those \acp{BCI} rely mostly on the P3 component, other components (\emph{e.g.}, occipital N1 and/or N200) may also be used for ERP detection \parencite{Bianchi2010, Kaufmann2011}, for this reason we prefer here to use the term \emph{oddball-based \acp{BCI}}.

@@

% SSVEP BCI, definition, plus and minus
Other systems of interest are \acp{BCI} based on \acp{SSVEP}.
They rely on the psychophysiological properties of the EEG brain responses recorded from the occipital cortex during the periodic presentation of identical visual stimuli (\emph{i.e.} flickering stimuli).
When the periodic presentation is at a sufficiently high rate (\SI{>6}{\Hz}), stable and synchronized neural oscillations at the stimulus frequency and its harmonics are evoked over the visual cortex \parencite{Regan1966, Herrmann2001, Luck2005}.
Such \acp{BCI} are particularly attractive because \acp{SSVEP} have high signal-to-noise ratios and are less susceptible to eye movement and blink artifacts \parencite{Perlstein2003} as well as electromyographic artifacts \parencite{Gray2003}.
Several \ac{SSVEP}-based \acp{BCI} have been successfully tested with healthy subjects (see \cite{Vialatte2010} for a review).

@@

    % Stimuli presentation
    All stimulation employed MATLAB\textsuperscript{\textregistered}, the stimuli were visually presented on a laptop's LCD screen (\SI{60}{\Hz} refresh rate) and their display and timing used the \emph{Psychophysics Toolbox Extensions} \parencite{Brainard1997,Pelli1997}.

@@

        % exp description
        As shown in \autoref{fig:stimSeq}, a typical \emph{stimulation cycle}, started with a \SI{2000}{\ms} cue, indicating the participant his/her target item, followed by a \SI{1000}{\ms} pause during which the cue disappeared and all icons remained gray.
        The background rectangle started then to flicker and the oddball stimulation began \SI{500}{\ms} later.
        The oddball stimulation consisted of 10~\emph{flashing sequences} during which each of the 6~icons was flashed one after another in random order for a duration randomly set between \SIlist[list-units = single]{200;300}{\ms}.
        As usually done for oddball experiments, the participants were instructed to focus on their target symbol and count the number of time it flashes.
        A \SI{1000}{\ms} pause followed the oddball stimulation and preceded the next cue.
        An \emph{experimental run} lasted approximately 4~minutes and consisted of 12 consecutive stimulation cycles, so that each of the 6~icons was cued twice (in random order).

        As we aimed here at studying the effect of the flickering background on the oddball \ac{ERP} response, we considered 5~experimental conditions.
        The first one (\emph{baseline condition}) consisted of a run as described in the previous paragraph but in which no flickering background was displayed.
        The 4~other conditions (\emph{hybrid conditions}) differed only by the frequency of the flickering background; the frequencies used were \SIlist[list-units = single]{8.57;10;12;15}{\Hz}, corresponding to the division of the refreshing rate of the screen by \numlist{7;6;5;4}, respectively.

@@

        % use
%        The data collected were used to compare the shape of the oddball response (response to the flashing of the target icon) and the \ac{ERP} classification accuracy (response to target \emph{v.s.} response to non-target flasing) across conditions.

@@

        % use
%        This experimental design allows a comparison of the \ac{SSVEP} responses of the participants with and without an oddball stimulation superimposed on the \ac{SSVEP} stimulation for a different set of \ac{SSVEP} frequencies.

@@

%        After studying individually oddball \acp{ERP} and \ac{SSVEP} responses, we aim here at studying the possibility to detect both responses
        This third experiment consists in a proof-of-concept for a hybrid oddball-\ac{SSVEP} \ac{BCI}.
        N subjects took part in the experiment.

        Two rectangles flickering at \SI{12}{\Hz} and \SI{15}{\Hz} where simultaneously presented on the left and right side of the screen, respectively.
        Within each of those rectangles 6~items were presented so that 2~independent and simultaneous oddball paradigm could occur as shown in \autoref{fig:stim2oddball}.
        The \emph{stimulation cycle} was the same as described in \autoref{sec:2.2.1ProtocolOddball}, icons from the left and right rectangles were always flashed simultaneously, however the order in which the icons would be flashed was set independently (and randomly) for each rectangle.
        An \emph{experimental run} lasted approximately 4~minutes and consisted of 12 consecutive stimulation cycles, so that each of the 12~icons was cued once (in random order).
        Each subject participated in 8~consecutive runs with 5~to 10~minutes pause after the the 4\nth run.

@@

        \subsubsection{Experiment 1: \acs{ERP} classification}
        \label{sec:2.3.1AnalysisExp1}

        % observe ERP responses
        We first observed average responses to target (\hl{and non-target?}) stimuli for each of the 5~experimental conditions.
        The \ac{EEG} signals were filtered between \SIlist[list-units = single]{0.3;30}{\Hz} (zero-phase 3\rd order Butterworth filter) and epochs were cut from \SI{200}{\ms} before the stimuli onsets until \SI{800}{\ms} after.
        In order to ensure that none of the epochs used for averaging were corrupted by ocular artifact, we rejected, for each \emph{experimental run}, the 15\% epochs with the highest peak-to-peak amplitude \parencite{Luck2005}.
        We also visually inspected the filtered \ac{EEG} traces to verify that no of ocular artifact could be seen within the 85\% remaining epochs.
        For each participant, averaged \acp{ERP} were observed and compared with respect to the experimental condition.
        We particularly looked for differences between the baseline condition (pure oddball) and the hybrid conditions (4 other condition with flickering square) and within the hybrid conditions themselves for an eventual influence of the flickering frequency over the \ac{ERP} response.

        % classify the ERPs
        The second step was to compare classification accuracies.
        The \ac{EEG} signals were filtered between \SIlist[list-units = single]{0.5;20}{\Hz} (zero-phase 3\rd order Butterworth filter), epochs were cut from each the stimuli onsets until \SI{600}{\ms} after and downsampled to \SI{128}{\Hz}.
        The resulting epochs were labeled to either \emph{target epochs} or \emph{non-target epochs} according to whether they corresponded to the \ac{EEG} response to a target stimulus (flashing of a target symbol) or a non-target one (flashing of any non-target symbol).
        For each subject and experimental condition, we ran a 3-fold cross-validation where a linear \ac{SVM} was trained \parencite{Keerthi2006} on the data collected during 2~out of the 3~experimental runs and the performance was measured on the remaining run.
        We thus obtain for each subject and experimental condition 36~correctness values (0: wrongly detected and 1: correctly detected).
        The correctness values were computed for a number of repetitions $N_r$ of the flashing sequence varying from 1~to 10.
        In order to mimic the behavior of a \ac{BCI}, for each stimulation cycle and each icon, epochs were average over the \emph{$N_r$ first repetitions}.

        % statistical model
        The correctness data were analysed using R (\hl{CITE}) and the R package \emph{lme4} (\hl{CITE + languageR?}).
        We used logistic linear mixed effect models (\hl{CITE}) with the number of repetitions nested within subjects as random factors.
        As fixed factors, we considered the experimental condition, the number of repetitions and the interaction between those 2~factors.
        The significance of the fixed factors as predictors for the correctness was established by means of likelihood ratio test (\hl{CITE}).

%%%%%%        % preprocessing: filtering
%%%%%%
%%%%%%        % studying the shape of ERP: epoch rejection
%%%%%%
%%%%%%        % studying classification accuracy: no epoch rejection
%%%%%%        We are interested in studying the effect of the flickering stimulus on the target \ac{ERP} detection accuracy.
%%%%%%        As the accuracy of an oddball \ac{BCI} is known to depend on the number of repetitions of flashing sequences used to average the \acp{ERP}; the classification accuracy was measured on \acp{ERP} averaged over the $n$~first repetitions of the flashing sequence with $n$~going from 1~to~10.
%%%%%%        This was done so as to mimic a real use of the system (where the stimulation sequence would stop after $n$~repetitions) and so that the accuracies are always measured on the same number of detection attempts.
%%%%%%
%%%%%%        For each experimental condition
%%%%%%
%%%%%%
%%%%%%        Data were filtered between \SIlist[list-units = single]{0.3;15}{\Hz} (zero-phase 3\rd order Butterworth filter), cut into \SI{800}{\ms} epochs starting from the stimuli onsets and downsampled to \SI{128}{\Hz}.
%%%%%%
%%%%%%        For each trial (stimulus), we had 8~channels~$\times$~80~data~points~=~640~features to classify as a response to either a target stimulus or a non-target stimulus.
%%%%%%        A linear \emph{Support Vector Machine} (SVM; \cite{Cristianini2000,Suykens2002}) with a 10-fold cross-validation and a linesearch for the optimization of the regularization parameter was built from the normalized training features.
%%%%%%        Training the linear SVM with the modified finite Newton method proposed by \textcite{Keerthi2006} took around one minute.

@@

    \subsection{Experiment 1: studying the oddball \acsp{ERP}}
    \label{sec:3.1Oddball}

    \subsection{Experiment 2: studying the \acs{SSVEP} responses}
    \label{sec:3.2SSVEP}

    \subsection{Experiment 3: hybrid classification}
    \label{sec:3.3Hybrid}
